---
title: 'Example: data'
author: "Thijs Janzen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Example: data}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
required <- c("ape", "pheatmap", "phytools")
if (!all(unlist(lapply(required,
                       function(pkg) requireNamespace(pkg, quietly = TRUE)))))
  knitr::opts_chunk$set(eval = FALSE)

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6)
knitr::opts_chunk$set(fig.height = 6)
```

## Data

The treestats package can rapidly calculate summary statistics on
phylogenetic trees, and in this vignette, we demonstrate this on
empirical trees. We will make use of family-level pruned trees stemming
from the clootl supertree of birds. These were created for the original
publication accompanying the treestats paper.

```{r read_trees}
focal_trees <- ape::read.tree(file = "https://raw.githubusercontent.com/thijsjanzen/treestats-scripts/main/datasets/phylogenies/fracced/birds.trees")  # nolint
```

We can now calculate all summary statistics for all trees:

```{r calc_sum_stats}
all_stats <- c()
for (i in seq_along(focal_trees)) {
  if (!ape::is.ultrametric(focal_trees[[i]])) {
    testthat::expect_output(
    focal_trees[[i]] <- phytools::force.ultrametric(focal_trees[[i]])
    )
  }
  focal_stats <- treestats::calc_all_stats(focal_trees[[i]])
  all_stats <- rbind(all_stats, focal_stats)
}
all_stats <- as.data.frame(all_stats)
```

We can now, for instance, plot the distribution of family sizes in
birds:

```{r plot_size}
hist(all_stats$number_of_lineages, main = "Number of lineages",
     xlab = "Number of lineages", ylab = "count")
```

Furthermore, we can make a heatmap of all correlations:

```{r corr}
cor_dist <- cor(all_stats)
diag(cor_dist) <- 0
heatmap(cor_dist)
```

This will generate a distorted image: correlations are not corrected for
tree size. We can study this a bit more in detail:

```{r plot with tree size}
opar <- par()
par(mfrow = c(3, 3))
for (stat in c("area_per_pair", "colless", "eigen_centrality",
               "four_prong", "max_betweenness", "max_width",
               "mntd", "sackin", "wiener")) {
  if (stat != "number_of_lineages") {
    x <- all_stats[, colnames(all_stats) == "number_of_lineages"]
    y <- all_stats[, colnames(all_stats) == stat]
    plot(y ~ x, xlab = "Tree size", ylab = stat, pch = 16)
  }
}
par(opar)
```

To correct for this, we will have to go over the entire correlation
matrix.

```{r correct_corr}
tree_size <- all_stats[, colnames(all_stats) == "number_of_lineages"]

for (i in seq_len(nrow(cor_dist))) {
  for (j in seq_len(ncol(cor_dist))) {
    stat1 <- rownames(cor_dist)[i]
    stat2 <- colnames(cor_dist)[j]
    x <- all_stats[, colnames(all_stats) == stat1]
    y <- all_stats[, colnames(all_stats) == stat2]

    a1 <- lm(x ~ tree_size)
    a2 <- lm(y ~ tree_size)
    new_cor <- cor(a1$residuals, a2$residuals)
    cor_dist[i, j] <- new_cor
  }
}
diag(cor_dist) <- 0
heatmap(cor_dist)
```

A nicer way to visualize this is given by the package ppheatmap:

```{r nicer plot, out.width="100%" }
if (requireNamespace("pheatmap")) pheatmap::pheatmap(cor_dist)
```

## Adding a random category to our matrix

To test whether outliers are perhaps instead of identifying unique
information, picking up on random noise, we can also add an extra
statistic that has no meaning, and that has random correlations with all
other statistics:

```{r add random}
cor_dist2 <- cor_dist
rand_values <- runif(n = nrow(cor_dist2))
cor_dist2 <- cbind(cor_dist2, rand_values)
cor_dist2 <- rbind(cor_dist2, c(rand_values, 0))
colnames(cor_dist2)[ncol(cor_dist2)] <- "random"
rownames(cor_dist2)[nrow(cor_dist2)] <- "random"

# and now we plot again:
heatmap(cor_dist2)
if (requireNamespace("pheatmap")) pheatmap::pheatmap(cor_dist2)
```
