---
title: "Getting started"
author: "Thijs Janzen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.width = 6)
knitr::opts_chunk$set(fig.height = 6)
library(treestats)
```

# Using treestats

The treestats package provides an easy to use interface to calculate summary
statistics on phylogenetic trees. To obtain a list of all supported summary 
statistics use:

```{r list}
list_statistics()
```

If your favourite summary statistic is missing, please let the maintainer know,
treestats is a dynamic package always under development, and the maintainers are
always looking for new statistics!

Given a phylogenetic tree, you can now use of the available functions to
calculate your summary statistic of choice. Let's take for instance the Colless
statistic (and we generate a dummy tree):

```{r colless}
phy <- ape::rphylo(n = 100, birth = 1, death = 0.1)

treestats::colless(phy)
```

Looking at the documentation of the colless statistic (`?colless`), we find that
the function also includes options to normalize for size: either 'pda' or
'yule':
```{r colless2}
treestats::colless(phy, normalization = "yule")
```

## Multiple statistics
The treestats package supports calculating many statistics in one go. For this,
several functions have been set up aptly. Firstly, the function 
`calc_all_stats` will calculate all statistics:
```{r all_stats}
all_stats <- calc_all_stats(phy)
```

Similarly, we can also blanket apply all topology associated summary statistics:
```{r all_stats3}
balance_stats <- calc_topology_stats(phy)
unlist(balance_stats)
```
